---
title: "Penguins"
author: "Anabelle Marques Klovrza"
date: "2024-09-25"
output: 
html_document:
    highlight: tango
    number_sections: no
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
urlcolor: blue
linkcolor: purple
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
header-includes: \linespread{1.05}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=F,
                      fig.path = 'figs/',
                      cache.path='cache/',
                      warning=F,
                      message=F)

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```

# Discussion 5 - Making a pipeline in group
We will do the `penguins`. First, we filtered the data from `penguins_raw` and grouped it by `Island`and `Species`:
```{r}
library(tidyverse)
library(palmerpenguins)

penguins_by_island_species <- penguins_raw |> 
    drop_na() |>
    select(!Comments) |>
  group_by(Island, Species) |>
    arrange(desc(`Body Mass (g)`))
```

Now, we want to see specific islands:
```{r}
biscoe <- penguins_by_island_species |>
  filter(Island == "Biscoe")

view(biscoe)

torgersen <- penguins_by_island_species |>
  filter(Island == "Torgersen")

view(torgersen)

dream <- penguins_by_island_species |>
  filter(Island == "Dream")

view(dream)
```

# PCA of Antarctic penguins 
Packages to create the PCA:
```{r}
library(tidyverse)
library(ggplot2)
library(tidyr)
library(purrr)
library(factoextra)
library(plyr)
library(dplyr)
library(vegan)
library(ggpubr)
library(RColorBrewer)
library(scales)
library(car)
library(devtools)
library(ggbiplot)
library(ggfortify)
library(pairwiseAdonis)
```

According to the variables `Culmen Length (mm)`, `Culmen Depth (mm)`, `Body Mass (g)`, and `Flipper Length (mm)`. We can see that there are two main clusters: a Gentoo cluster, and a Adelie-Chinstrap cluster!
```{r PCA}
pcapenguins <- prcomp(penguins_by_island_species[, c("Body Mass (g)", "Culmen Length (mm)", "Culmen Depth (mm)", "Flipper Length (mm)" )], 
                     center = TRUE, 
                     scale. = TRUE)
pcapenguins

summary(pcapenguins)

# Variance 
variance_pca = (pcapenguins$sdev)^2

variance_pca

# extract autovalues - proportion of variance on the PCs
eig.val <- get_eigenvalue(pcapenguins)
eig.val

# Variables
res.var <- get_pca_var(pcapenguins)
res.var$coord

## contribution for the PCs - each variable contributes in percentage
res.var$contrib
res.var$cos2 # representation quality

# correlation plot
library(corrplot)
corrplot(res.var$cos2, is.corr=FALSE)

# variance proportions for each variable
fviz_eig(pcapenguins, addlabels = T)

# data frame with the PCA results
pca_data <- as.data.frame(pcapenguins$x)
pca_data$Species <- penguins_by_island_species$Species
pca_data$Island <- penguins_by_island_species$Island

# colors for each group
colorspenguins <- c("Adelie Penguin (Pygoscelis adeliae)" = "darkolivegreen2", "Gentoo penguin (Pygoscelis papua)" = "skyblue1", "Chinstrap penguin (Pygoscelis antarctica)" = "firebrick1")

# graph
penguinsPCA <- ggplot(pca_data, aes(x = PC1, y = PC2), color = Species) +
  geom_point(aes(shape = Island, fill = Species)) +
  stat_ellipse(aes(color = Species), alpha = 1, size = 0.75, level = 0.90) + # add ellipse
  labs(title = "PCA of Antarctic penguins", subtitle = "For body mass, culmen length and depth, and flipper length", x = "PC1 (55.2%)", y = "PC2 (26.82%)") +
  theme_minimal() +
  theme(axis.title = element_text(color = "black"),
        axis.text = element_text(color = "black"),
        axis.line = element_line(color = "black")) +
  scale_color_manual(values = colorspenguins)

print(penguinsPCA)
  
# adding variables
loadings <- as.data.frame(pcapenguins$rotation) * sqrt(pcapenguins$sdev)

# adding text and arrows for the variables
penguin <- penguinsPCA + geom_segment(data = data.frame(xend = loadings[, 1], yend = loadings[, 2]),
                      aes(x = 0, y = 0, xend = xend, yend = yend), 
                      color = 'black', arrow = arrow(length = unit(0.09, "inches")), 
                      size = 1) +
          geom_text(data = data.frame(x = loadings[, 1], y = loadings[, 2], label = rownames(loadings)),
                    aes(x = x, y = y, label = label), 
                    color = 'black', size = 3.5, vjust = -1)

print(penguin)
```
